Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-04-08 WholePlotTrajectories.tex
review Bruno

\Authors{
Ariane MIRABEL\textsuperscript{1*}\\ Bruno Herault\textsuperscript{2}\\ Eric Marcon\textsuperscript{1}
} % Authors
\affiliation{
\textsuperscript{1}UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles,
Université de Guyane.\\ \hspace{1em} Campus Agronomique, 97310 Kourou, France.\\\textsuperscript{2}INPHB, Institut National Polytechnique Félix Houphoüet-Boigny\\ \hspace{1em} Yamoussoukro, Ivory Coast.
}
\affiliation{*\textbf{Corresponding author}: ariane.mirabel@ecofog.gf, https://github.com/ArianeMirabel} % Corresponding author

@@

The experiment is a network of twelve 6.25 ha plots (Table
\ref{tab:Tab1}) that underwent three disturbance treatments in 1987
according to a randomized plot design \citep{Gourlet-Fleury2004}.\\
The experiment comprised three replicates of three sylvicultural
treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with
50 cm minimum DBH that belonged to a set of 58 commercial species
\citep{Gourlet-Fleury2004}. Treatment T2 additionally comprised a
thinning treatment by poison-girdling of non-commercial, randomly
selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40
cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm
minimum DBH, all belonging to non-commercial species. Considering the
silvicultural treatments and the following damage, disturbance intensity
was measured as the percentage of aboveground biomass (\%AGB) lost
between the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Biomass2018}. The three treatments constituted a disturbance
intensity gradient with increasing of above-ground biomass (AGB) lost
and surface disturbed.

@@

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control & - & - & - & 0\\
T1, low & DBH $\geq$ 50 cm, commercial species, $\approx$ 10   $trees.ha^{-1}$ & - & - & $[12-33]$\\
T2, intermediate & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30   $trees.ha^{-1}$ & - & $[33-56]$\\
T3, high & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15  $trees.ha^{-1}$ & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species,\ $\approx$ 15 $trees.ha^{-1}$ & $[35-56]$\\
\bottomrule
\end{tabu}
\end{table}

@@

Six functional traits representing leaf economics (leaf thickness,
toughness, total chlorophyll content and specific leaf area) and stem
economics (wood specific gravity and bark thickness), and life-history
traits (maximum specific height and seed mass) came from the BRIDGE
project \footnote{http://www.ecofog.gf/Bridge/}. Trait values were
assessed from a selection of individuals located in nine permanent plots
in French Guiana, including two in Paracou, and comprised 294 species
pertaining to 157 genera. Missing trait values (10\%) were filled using
multivariate imputation by chained equation \citep{Mice2011}.
Imputations were restricted within genus or family when samples were too
scarce, in order to account for the phylogenetic signal. Whenever a
species was not in the dataset, it was attributed a set of trait values
randomly sampled among species of the next higher taxonomic level (same
genus or family). As seed mass information was classified into classes,
no data filling process was applied and analyses were restricted to the
414 botanical species recorded.

@@

Because of the variability in the precision of botanical identification
efforts, we were constrained to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination plane. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference inventories in 1989, \emph{i.e.} 2 years
after disturbance. Univariate trajectories of the leaf, stem and
life-history traits were also visualized with the community weighted
means (CWM) \citep{Diaz2007}. Species seed mass were given in 5 mass
classes, and seed mass trajectories were reported as the proportion of
each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the
Hill number translation of the Simpson index \citep{Hill1973}. These
metrics allowed assessing the taxonomic richness as well as evenness,
through the comparison between these two metrics.: results will be
discussed directly in terms of taxonomic richness and evenness. Both
indices belong to the set of HCDT or generalized entropy, respectively
corresponding to the 0 and 2 order of diversity (q), recommended for
diversity studies \citep{Marcon2015b}. The functional diversity was
reported using the functional richness and functional evenness,
\emph{i.e} the Rao index of quadratic entropy. The Rao index combines
species abundance distribution, and the average pairwise functional
dissimilarity between species computed by the gower distance.

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic richness \textbf{(a)}, Simpson diversity \textbf{(b)}, functional richness \textbf{(c)}, and Rao diversity \textbf{(d)}. Shaded areas are the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

@@

Before disturbance, the different communities had different taxonomic
compositions, visualized by their distinct starting points on the NMDS
axis 2. These initial differences were maintained all along the 30 years
following disturbance, with the disturbance leading a displacement on
the NMDS axis 1. The taxonomic compositional changes following
disturbance were similar among communities, which may correspond to the
recruitment of a group of pioneers common to all communities whenever
their initial differences and the intensity of disturbance
\citep{Denslow2000, Bongers2009}. Taxonomic trajectories continued with
a recovery of the initial composition which, althought not fully
achieved after 30 years, suggested the resilience of the taxonomic
composition resilience and the maintenance of the initial differences
among communities \citep{Folke2006}. The return towards initial
taxonomic composition showed that species not belonging to the
pre-disturbance community were rarely recruited, probably because of the
common dispersal limitations among tropical tree species
\citep{Svenning2005}.

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
at which the taxonomic richness was maximized and the taxonomic evenness
remained resilient. The disturbance intensity impacted the balance in
the community between pre-disturbance surviving trees, and trees
recruited afterward. On the one hand, below a 20\%-25\% AGB loss, the
trees surviving after disturbance remained numerous enough to maintain
the high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pioneers, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. On the other hand, above the
intensity threshold, the taxonomic richness of surviving trees was too
low to be offset by the recruitment of pioneers. In the Guiana shield
indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The taxonomic richness
then decreased following intense disturbance, all the more so that the
disturbance intensity was high \citep{Molino2001}. At the same time,
pioneers became persistently dominant and prevented a return towards the
initial evenness and the recovery of hyper-dominant shade-tolerant
species.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying an easy recruitment for the first species
but becoming increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The recovery of the functional
redundancy then relied upon the random process of species recruitment
and was increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

@@

Community post-disturbance trajectories in compositon and diversity were
shaped by the recruitment of a determined pool of pioneers identical
among communities and disturbance intensity. Nevertheless, taxonomic and
functional composition trajectories appeared decoupled. While functional
trajectories remained similar in the functional space and converged
towards the recovery of a comparable initial state, taxonomic
trajectories showed initial differences in community composition that
were maintained along time. This decoupling was explained by a decrease
of the functional redundancy that mitigated the functional impact of
disturbance. Community functional and taxonomic diversity response to
disturbance was contrasted as well. While the functional trajectories
remained similar whenever the disturbance intensity, taxonomic
trajectories were markedly different above a threshold of 20-25\% AGB
removed for which taxonomic richness was maximized and taxonomic eveness
remained resilient. The Intermediate Disturbance Hypothesis applied to
the linkage of disturbance with the taxonomic diversity, but not with
the functional diversity. Taxonomic diversity was maximized following a
disturbance of intermediate intensity, that was the threshold of 25\%
AGB removed, and at an intermediate time after disturbance, that was
around 25 years after disturbance when the recruitment of early- and
late-successional species were balanced. Whenever the disturbance
intensity, community resilience (in terms of recovery of the
pre-disturbance state) was tangible but required several decades and
relied upon the random lottery recruitment of rare species. Although
resilient, the functional redundancy was lower for more than 30 years
after disturbance which cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Herault2018}.

>>>>

\Authors{
Ariane MIRABEL\textsuperscript{1*}\\ Bruno Herault\textsuperscript{2}\\ Eric Marcon\textsuperscript{1}
} % Authors
\affiliation{
\textsuperscript{1}UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles,
Université de Guyane.\\ \hspace{1em} Campus Agronomique, 97310 Kourou, France.\\\textsuperscript{2}INPHB, Institut National Polytechnique Félix Houphoüet-Boigny\\ \hspace{1em} Yamoussoukro, Ivory Coast.
}
\affiliation{*\textbf{Corresponding author}: ariane.mirabel@gmail.com, https://github.com/ArianeMirabel} % Corresponding author

@@

The experiment is a network of twelve 6.25 ha plots (Table
\ref{tab:Tab1}) that underwent three disturbance treatments in 1987
according to a randomized plot design \citep{Gourlet-Fleury2004}.\\
The experiment comprised three replicates of three sylvicultural
treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with
50 cm minimum DBH that belonged to a set of 58 commercial species
\citep{Gourlet-Fleury2004}. Treatment T2 additionally comprised a
thinning treatment by poison-girdling of non-commercial, randomly
selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40
cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm
minimum DBH, all belonging to non-commercial species. Considering the
silvicultural treatments and the following damage, disturbance intensity
was measured as the percentage of aboveground biomass (\%AGB) lost
between the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Rejou2017}. The three treatments constituted a disturbance
intensity gradient with increasing of above-ground biomass (AGB) lost
and surface disturbed.

@@

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou. Treatment intensities are defined by the minimum logging DBH (Diameter at Breast Height), the type of logged species (commercial or not), the density of logged trees, and the total AGB (Above Ground Biomass) lost after treatment.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control & - & - & - & 0\\
T1, low & DBH $\geq$ 50 cm, commercial species, $\approx$ 10   $trees.ha^{-1}$ & - & - & $[12-33]$\\
T2, intermediate & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30   $trees.ha^{-1}$ & - & $[33-56]$\\
T3, high & DBH $\geq$ 50 cm, commercial species, $\approx$ 10  $trees.ha^{-1}$ & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15  $trees.ha^{-1}$ & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species,\ $\approx$ 15 $trees.ha^{-1}$ & $[35-56]$\\
\bottomrule
\end{tabu}
\end{table}

@@

Six functional traits representing leaf economics (leaf thickness,
toughness, total chlorophyll content and specific leaf area) and stem
economics (wood specific gravity and bark thickness) came from the
BRIDGE project \footnote{http://www.ecofog.gf/Bridge/}. Trait values
were assessed from a selection of individuals located in nine permanent
plots in French Guiana, including two in Paracou, and comprised 294
species pertaining to 157 genera. Whenever a species was in the dataset
but missed some trait values (10\% of the species), missing values were
filled using multivariate imputation by chained equation
\citep{Mice2011}. To account for the phylogenetic signal in the filling
process, imputations based on samples from the same genus or from the
same family. Whenever a species was not in the dataset, it was
attributed a set of trait values randomly sampled among species of the
next higher taxonomic level (same genus or family). Two life-history
traits (maximum specific height and seed mass) came from the Mariwenn
database \footnote{https://www.ecofog.gf/mariwenn/}. The database
compiles information from a vast literature on the flora of French
Guiana \citep{Ollivier2007} and comprises 362 species pertaining to 188
genera. As seed mass information was classified into classes, no data
filling process was applied and analyses were restricted to the
botanical species recorded.

@@

Because of the variability in the precision of botanical identification
efforts, we were constrained to conduct the taxonomic composition and
diversity analysis at the genus level. Taxonomic and functional
trajectories of community composition were followed in a two-dimensional
NMDS ordination plane. Two NMDS using abundance-based (Bray-Curtis)
dissimilarity measures were conducted to map either taxonomic or
functional composition, the latter based on the seven leaf, stem and
life history traits (without seed mass classes). Trajectories along time
were reported through the Euclidean distance between the target
inventories and the reference pre-disturbance inventories of 1984.
Univariate trajectories of the leaf, stem and life-history traits were
also visualized with the community weighted means (CWM)
\citep{Diaz2007}. Species seed mass were given in 5 mass classes, and
seed mass trajectories were reported as the proportion of each class in
the inventories (Supplementary materials).

The taxonomic and functional trajectories were analysed from the
reference pre-disturbance inventories of 1984. The taxonomic diversity
was reported through species richness and the Hill number translation of
the Simpson index \citep{Hill1973}. These metrics allowed assessing the
taxonomic richness as well as evenness, through the comparison between
these two metrics: results will be discussed directly in terms of
taxonomic richness and evenness. Both indices belong to the set of HCDT
or generalized entropy, respectively corresponding to the 0 and 2 order
of diversity (q), recommended for diversity studies \citep{Marcon2015}.
The functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} the Rao index of quadratic entropy. The
Rao index combines species abundance distribution, and the average
pairwise functional dissimilarity between species computed by the gower
distance.

@@

\caption{Trajectories of community taxonomic richness \textbf{(a)}, Simpson diversity \textbf{(b)}, functional richness \textbf{(c)}, and Rao diversity \textbf{(d)}. Values correspond to the difference over 30 years of community diversity with the values of reference 1984 pre-disturbance inventories. Shaded areas are the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

@@

Before disturbance, the different communities had different taxonomic
compositions, visualized by their distinct starting points on the NMDS
axis 2. These initial differences were maintained all along the 30 years
following disturbance, with the disturbance leading a displacement on
the NMDS axis 1. The taxonomic compositional changes following
disturbance were similar among communities, which may correspond to the
recruitment of a group of pioneers, as \emph{Cecropia} spp.,
\emph{Vismia} spp., common to all communities whenever their initial
differences and the intensity of disturbance
\citep{Denslow2000, Bongers2009}. Taxonomic trajectories continued with
a recovery of the initial composition which, althought not fully
achieved after 30 years, suggested the resilience of the taxonomic
composition resilience and the maintenance of the initial differences
among communities \citep{Folke2006}. The return towards initial
taxonomic composition showed that species not belonging to the
pre-disturbance community were rarely recruited, probably because of the
common dispersal limitations among tropical tree species
\citep{Svenning2005}.

@@

Comparing community diversity along time after disturbance revealed
constrasting trajectories according to the distrubance intensity. Below
a disturbance intensity threshold, in the case of treatments 1 and 2,
the community taxonomic richness increased. In parallell, after a first
increase, the taxonomic eveness stabilized or strated to decrease after
30 years. Below the intensity threshold Contrastingly, above an
intensity threshold, in the case of treatment 3, the taxonomic richness
did not exceeded the initial value and the taxonomic evenness
continiously increased for 30 years without sign of return towards the
initial value. Below and above a threshold, corresponding to a
disturbance intensity between treatments 2 and 3, community taxonomic
richness and evenness changed markedly. A the disturbance intensity
threshold, the taxonomic richness was maximized and the taxonomic
evenness remained resilient. The disturbance intensity impacted the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. On the one hand, below a 20\%-25\% AGB loss,
the trees surviving after disturbance remained numerous enough to
maintain the high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pioneers, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. On the other hand, above the
intensity threshold, the taxonomic richness of surviving trees was too
low to be offset by the recruitment of pioneers. In the Guiana shield
indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The taxonomic richness
then decreased following intense disturbance, all the more so that the
disturbance intensity was high \citep{Molino2001}. At the same time,
pioneers became persistently dominant and prevented a return towards the
initial evenness and the recovery of hyper-dominant shade-tolerant
species.

@@

The decoupling between taxonomic and functional trajectories was
accompanied by a decrease in the functional redundancy within the
pre-disturbance functional space. This decrease was explained by a loss
of species following disturbance and then a lower redundancy of their
functional characteristics. This decrease of the functional redundancy
was not compensated by the recruitment of pioneers, as those were
functionally different from the pre-disturbance community. Progressively
though, more competitive,long-lived pioneers or late-successional
species functionally closer to the pre-disturbance community
progressively replaced the first established species and restored the
functional redundancy. This replacement followed the lottery recruitment
rules, implying an easy recruitment for the first species but becoming
increasingly hampered by the emergence of interspecific competition
\citep{Busing2002}. The recovery of the functional redundancy then
relied upon the random process of species recruitment and was
increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

@@

Community post-disturbance trajectories in compositon and diversity were
shaped by the recruitment of a determined pool of pioneers identical
among communities and disturbance intensity. Nevertheless, taxonomic and
functional composition trajectories appeared decoupled. While functional
trajectories remained similar in the functional space and converged
towards the recovery of a comparable initial state, taxonomic
trajectories showed initial differences in community composition that
were maintained along time. Community high functional redundancy could
explain this decoupling as the loss of a species does not necessarily
the loss of its functional charactristics in the community. In this
sense, the functional redundancy would mitigate the functional impact of
disturbance. Community functional and taxonomic diversity response to
disturbance were contrasted as well. While the functional trajectories
remained similar whenever the disturbance intensity, taxonomic
trajectories were markedly different above a threshold of 20-25\% AGB
removed for which taxonomic richness was maximized and taxonomic eveness
remained resilient. The Intermediate Disturbance Hypothesis applied to
the linkage of disturbance with the taxonomic diversity, but not with
the functional diversity. Taxonomic diversity was maximized following a
disturbance of intermediate intensity, that was the threshold of 25\%
AGB removed, and at an intermediate time after disturbance, that was
around 25 years after disturbance when the recruitment of early- and
late-successional species were balanced. Whenever the disturbance
intensity, community resilience (in terms of recovery of the
pre-disturbance state) was tangible but required several decades and
relied upon the random lottery recruitment of rare species. Although
resilient, the functional redundancy was lower for more than 30 years
after disturbance which cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Herault2018}. Our conclusions first highlighted risks to lose
rare species with visible impacts only in the long term, and led us to
advice paying particular attention to rare and undangered species and to
conduct regular inventories during exploitation times. Second,
post-disturbance trajectories seemed to greatly rely on post-disturbance
recruitment dynamics that would be key to model post-disturbance
trajectories.
